Detection of Oscillations in Aldebaran with Ground-Based Observations

### 2018-01-18 paper/aldebaran.tex
carma appendix; placeholder text in body

A Gaussian Process such as that used by \citet{brewer2009} also corresponds to
the solution, $y(t)$, of the stochastic ODE \begin{equation}
\label{eq:GP-definition} \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
\prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t), \end{equation} where
$p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process; $r_i \in
\mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation of the
ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the inverse
process; and $\eta(t)$ is a white-noise Gaussian Process with $\left\langle
\eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t') \right\rangle =
\sigma^2 \delta\left(t - t' \right)$.  Such a solution is called a CARMA
(Continuous AutoRegressive Moving Average) process.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is \begin{equation}
\label{eq:CARMA-PSD} P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i
f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation} The power spectrum is a rational function of frequency with poles
at $2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the invertable
constraints, these poles and zeros all occur when $\Im f \neq 0$.  The
autocorrelation function corresponding to the power spectrum in Eq.\
\eqref{eq:CARMA-PSD} is \begin{equation} \label{eq:CARMA-ACF} \rho_y (\tau) =
\left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle =
\sum_{i=1}^p A_i^2 e^{r_i \tau}, \end{equation} where the RMS amplitudes of the
$p$ different modes, $A_i$, are functions of $\sigma$ and the values of the
roots $r_i$ and $b_j$.  If $q = p-1$, then the $A_i$ are independent, while if
$q < p-1$ then there are interdependencies among the $A_i$.  From either the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function in
Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$ represents an
exponentially-decaying mode with $r_i$ the rate constant, while a complex
conjugate pair of $r_i$ represent an oscillatory mode with decay rate $\Re r_i$
and angular frequency $\left|\Im r_i\right|$.  Oscillatory modes can also be
described by their frequency $f$ and quality factor $Q$ via $r = f/(2Q) \pm 2\pi
i f$.  Decaying modes generate a damped random walk, also known as a
Ornstein-Uhlenbeck process.  \citet{Kelly2014} gives more details on CARMA
processes in an astronomical context, including formulae for $A_i$ as a function
of $\sigma$, $r_i$, and $b_j$, and shows how a state-space model of the solution
$y(t)$ can be tracked through a time-series of uncertain observations $y_k =
y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of zero mean
and known (heteroskedastic) variance, using a Kalman filter to produce a
likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be used to
reduce the standard Gaussian Process covariance matrix for the process with
autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under observations at times
$t_k$ to a banded form that can then be Cholesky-decomposed in the standard
Gaussian Process likelihood function in $\mathcal{O}\left( n \right)$ time.

@@

>>>>

Fortunately, this GP can be written as the solution of a class of stochastic ordinary differential equations, the Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can be evaluated in linear time. For problems such as the damped driven harmonic oscillator, we therefore have access to these powerful computational tools for inferring their power spectral densities. \citet{Kelly2014} gives more details on CARMA
processes in an astronomical context, and shows how a state-space model of the solution
$y(t)$ can be tracked through a time-series of uncertain observations $y_k =
y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of zero mean
and known (heteroskedastic) variance, using a Kalman filter to produce a
likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be used to
reduce the standard Gaussian Process covariance matrix for the process with
autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under observations at times
$t_k$ to a banded form that can then be Cholesky-decomposed in the standard
Gaussian Process likelihood function in $\mathcal{O}\left( n \right)$ time.

@@

\section{CARMA Models}
\label{carmatheory}

A Gaussian Process such as that used by \citet{brewer2009} also corresponds to
the solution, $y(t)$, of the stochastic ODE \begin{equation}
\label{eq:GP-definition} \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
\prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t), \end{equation} where
$p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process; $r_i \in
\mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation of the
ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the inverse
process; and $\eta(t)$ is a white-noise Gaussian Process with $\left\langle
\eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t') \right\rangle =
\sigma^2 \delta\left(t - t' \right)$.  Such a solution is called a CARMA
(Continuous AutoRegressive Moving Average) process.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is \begin{equation}
\label{eq:CARMA-PSD} P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i
f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation} The power spectrum is a rational function of frequency with poles
at $2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the invertable
constraints, these poles and zeros all occur when $\Im f \neq 0$.  The
autocorrelation function corresponding to the power spectrum in Eq.\
\eqref{eq:CARMA-PSD} is \begin{equation} \label{eq:CARMA-ACF} \rho_y (\tau) =
\left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle =
\sum_{i=1}^p A_i^2 e^{r_i \tau}, \end{equation} where the RMS amplitudes of the
$p$ different modes, $A_i$, are functions of $\sigma$ and the values of the
roots $r_i$ and $b_j$.  If $q = p-1$, then the $A_i$ are independent, while if
$q < p-1$ then there are interdependencies among the $A_i$.  From either the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function in
Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$ represents an
exponentially-decaying mode with $r_i$ the rate constant, while a complex
conjugate pair of $r_i$ represent an oscillatory mode with decay rate $\Re r_i$
and angular frequency $\left|\Im r_i\right|$.  Oscillatory modes can also be
described by their frequency $f$ and quality factor $Q$ via $r = f/(2Q) \pm 2\pi
i f$.  Decaying modes generate a damped random walk, also known as a
Ornstein-Uhlenbeck process.
